---
title: "Data Visualization"
author: "Camila Vargas Poulsen"
date: "2023-03-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

## Load libraries

```{r}
library(leaflet)
library(dplyr)
library(tidyr)
library(ggplot2)
library(DT)
library(scales)
```

## Read Data

Load the data table directly from the KNB Data Repository, if it isn’t already present on your local computer. This technique only downloads the file if you need it.

This is a different and more advanced way to read data. The `tryCatch()` function runs an expression, in this case `read.csv()` and if it gets an error, it catches this error and do somthing else. In this case it goes at get the data from the url.

```{r read data, warning=FALSE, message=FALSE}
data_url <- "https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Af119a05b-bbe7-4aea-93c6-85434dcb1c5e"

esc <- tryCatch(
    read.csv("data/escapement.csv"),
    error=function(cond) {
        message(paste("Escapement file does not seem to exist, so get it from the KNB."))
        esc <- read.csv(url(data_url, method = "libcurl"))
        return(esc)
    }
)

head(esc)
glimpse(esc)

```



##Challenge

Now that we have the data loaded, use your `dplyr` and `tidyr` skills to calculate annual escapement by species and region.

Hint: try to use `separate` to extract the year, month, and day from the date column.


## Summary Data

Calculate the annual escapement by species and region.

```{r summary data}
annual_esc <- esc %>% 
  separate(sampleDate, c("Year", "Month", "Day"), sep = "-") %>% 
  mutate(Year = as.numeric(Year)) %>% 
  group_by(Year, Species, SASAP.Region) %>% 
  summarise(escapement = sum(DailyCount)) %>% 
  filter(Species %in% c("Chinook", "Sockeye", "Chum", "Coho", "Pink"))

head(annual_esc)
```


The chunck above used a lot of the dplyr commands that we’ve used, and some that are new. 
- The `separate` function is used to divide the sampleDate column up into Year, Month, and Day columns, 
- `group_by` to indicate that we want to calculate our results for the unique combinations of species, region, and year. 
- We next use `summarize` to calculate an escapement value for each of these groups. 
- Finally, we use a `filter` and the %in% operator to select only the salmon species.


## Static Plots

Basic bar plot.

```{r}
ggplot(annual_esc,
       aes( x = Species, 
            y = escapement))+
  geom_col()


```


Basic plot with blue bars.

```{r}
## Create this error
ggplot(annual_esc,
       aes( x = Species, 
            y = escapement,
            fill = "blue"))+
  geom_col()
```

**What happen here.**
- Clearly we did not achieve our goal of getting blue bars.
- What happened is that because we used the fill argument inside out `aes()` function, under the hood, ggplot created a new variable to our data or a new column and with the value "blue". 
- Anytime we want `ggplot` to adjust somthing based on one of the variables of oour data we put that argument inside the `aes()` function. 
- However, because we just want to change the color of the bars based on something cosmetics, just make them blue. We need to provide that argument outside the `aes()` function.

**So lets try this again.**

```{r}
ggplot(annual_esc,
       aes( x = Species, 
            y = escapement))+
  geom_col(fill = "blue")
```


Mapping each species escapement by year.

**Comment:** This is what makes `ggplot` is really powerful because we can easily get this plot to visualize more aspects of our data.

```{r}
ggplot(annual_esc,
       aes(x = Species,
           y = escapement,
           fill = SASAP.Region))+
  geom_col()
```

## `ggplot2` and the pipe operator (%\>%)

Just like in `dplyr` and `tidyr`, we can also pipe a data.frame directly into the first argument of the ggplot function using the `%>%` operator.

Let’s look at an example using a different geometry. Here, we use the pipe operator to pass in a filtered version of annual_esc, and make a line plot with points at each observation.


Annual escapement for Kodiak region

```{r ggplot and pipe}
annual_esc %>% 
  filter(SASAP.Region == "Kodiak") %>% 
  ggplot(aes(x = Year,
             y = escapement,
             color = Species))+
  geom_line()+
  geom_point()
  
```

## Setting themes in `ggplot`


- We have a plot, is looking good, but there are still things we can make better. For example rename the Y axis and the legend, maybe give my plot a title, remove the x axis label? Change the background.

- `ggplot` allows you to do a lot of customization. That is what is so creat about this tool

- We are going to use the `lab()` function to modify all the labels and the `theme_bw` function to change the overall aspect of out plot.

- For clarity in the next section, I’ll save the filtered version of the annual escapement data.frame to it’s own object.


Filtering data for Kodiak region
```{r kodiak}
kodiak_esc <- annual_esc %>% 
  filter(SASAP.Region == "Kodiak")
```


Plotting Kodiak data using `theme_`

```{r}
ggplot(kodiak_esc,
       aes(x = Year,
           y = escapement,
           color = Species))+
  geom_line()+
  geom_point()+
  ylab("Escapement")+
  ggtitle("Kodial Salmon Escapement")+
  theme_bw()
```


**Challenge**

- Use the RStudio autocomplete, the ggplot2 documentation, a cheatsheet, or good old google to find other built in themes. Pick out your favorite one and add it to your plot.

---

The built in theme functions change the default settings for many elements that can also be changed individually using the `theme()` function. The `theme()` function is a way to further fine-tune the look of your plot. This function takes MANY arguments (just have a look at ?theme). Luckily there are many great ggplot resources online so we don’t have to remember all of these, just google “ggplot cheatsheet” and find one you like.

Let’s look at an example of a theme call, where we change the position of our plot above from the right side to the bottom, and remove the title from the legend.


Customizing plot with `theme()`

```{r}
ggplot(kodiak_esc,
       aes(x = Year,
           y = escapement,
           color = Species))+
  geom_line()+
  geom_point()+
  ylab("Escapement")+
  ggtitle("Kodial Salmon Escapement")+
  theme_bw()+
  theme(legend.position = "bottom",
        legend.title = element_blank())
```

- Note that the theme() call needs to come after any built in themes like theme_bw() are used. Otherwise, theme_bw() will likely override any theme elements that you changed using theme().

- You can also save the result of a series of theme() function calls to an object to use on multiple plots. This prevents needing to copy paste the same lines over and over again!

Saving my theme

```{r my theme}
my_theme <- theme_bw()+
  theme(legend.position = "bottom",
        legend.title = element_blank())
```


plotting calling my saved them

```{r}
ggplot(kodiak_esc,
       aes(x = Year,
           y = escapement,
           color = Species))+
  geom_line()+
  geom_point()+
  ylab("Escapement")+
  ggtitle("Kodiak Salmon Escapement")+
  my_theme

```

- Fixing tick labels in ggplot can be super annoying. The y-axis labels in the plot above don’t look great. We could manually fix them, but it would likely be tedious and error prone.

- The scales package provides some nice helper functions to easily rescale and relabel your plots. Here, we use scale_y_continuous from ggplot2, with the argument labels, which is assigned to the function name comma, from the scales package. This will format all of the labels on the y-axis of our plot with comma-formatted numbers.

Smarter tick labels using scales

```{r}
ggplot(kodiak_esc,
       aes(x = Year,
           y = escapement,
           color = Species))+
  geom_line()+
  geom_point()+
  scale_y_continuous(labels = comma)+
  ylab("Escapement")+
  ggtitle("Kodiak Salmon Escapement")+
  my_theme
```



## Saving my plot
- Saving plots using ggplot is easy! The ggsave function will save either the last plot you created, or any plot that you have saved to a variable. You can specify what output format you want, size, resolution, etc.

Using `ggsave()` to save my plot as a png

```{r}
ggsave("kodiak_esc.png",
       width = 12,
       height = 3,
       units = "in")
```


## Creating multiple plots
-   What if we wanted to generate a plot for every region?
-   A fast way to do this uses the function facet_wrap(). This function takes a mapping to a variable using the syntax ~variable_name. The ~ (tilde) is a model operator which tells facet_wrap to model each unique value within variable_name to a facet in the plot.

Using `facer_wrap()` to create one plot per region.

```{r}
ggplot(annual_esc,
       aes(x = Year,
           y = escapement,
           color = Species))+
  geom_line()+
  geom_point()+
  scale_y_continuous(labels = comma)+
  facet_wrap(~SASAP.Region,
             scales = "free_y",
             ncol = 2)+
  ylab("Escapement")+
  my_theme
```


## Interactive tables

- Now that we know how to make great static visualizations, lets introduce two other packages that allow us to display our data in interactive ways. These packages really shine when used with GitHub pages, so at the end of this lesson we will publish our figures to the website we created earlier

Create interactive table for locations with `DT` package.

Create a new data frame with all locations and their coordinates.
```{r locations df}
locations <- esc %>% 
  distinct(Location, Latitude, Longitude) %>% 
  drop_na()
```

Create an interactive table
```{r}
datatable(locations)
```


## Interactive Maps

-   Similar to ggplot2, you can make a basic leaflet map using just a couple lines of code. Note that unlike ggplot2, the leaflet package uses pipe operators (%>%) and not the additive operator (+).

-   The addTiles() function without arguments will add base tiles to your map from **OpenStreetMap**. 
-   addMarkers() will add a marker at each location specified by the latitude and longitude arguments. Note that the ~ symbol is used here to model the coordinates to the map (similar to facet_wrap in ggplot).

Plot locations in an interactive map using `leaflet`

Default OpenStreetMap
```{r}
leaflet(locations) %>% 
  addTiles() %>% 
  addMarkers(lng = ~Longitude,
             lat = ~Latitude,
             popup = ~Location)
```


Using General Bathymetric Map of the Ocean (GEBCO) Web Map Service (WMS).

```{r}
leaflet(locations) %>% 
  addWMSTiles("https://www.gebco.net/data_and_products/gebco_web_services/web_map_service/mapserv?request=getmap&service=wms&BBOX=-90,-180,90,360&crs=EPSG:4326&format=image/jpeg&layers=gebco_latest&width=1200&height=600&version=1.3.0",
              layers = 'GEBCO_LATEST',
              attribution = "Imagery reproduced from the GEBCO_2022 Grid, WMS 1.3.0 GetMap, www.gebco.net") %>%
  addCircleMarkers(lng = ~Longitude,
                   lat = ~Latitude,
                   popup = ~ Location,
                   radius = 5,
                   # set fill properties
                   fillColor = "salmon",
                   fillOpacity = 1,
                   # set stroke properties
                   stroke = T,
                   weight = 0.5,
                   color = "white",
                   opacity = 1)
```



